cardinalR: Generating interesting high-dimensional data structures

### 2025-09-19 paper-cardinalR.Rmd
Updated S-curve section

An S-curve structure (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:scurve)"} else { "Figure \\@ref(fig:scurve-proj)"})`) simulates data that lies along a smooth, non-linear manifold, making it a classic example for testing the performance of non-linear dimension reduction methods. This package provides functions to generate standard S-curve shapes as well as variants containing holes to introduce missing or incomplete regions. These variations are useful for evaluating how well algorithms capture non-linear geometry and handle incomplete manifolds in high-dimensional data. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:scurve-tb-html)', '\\@ref(tab:scurve-tb-pdf)'))` summarizes these functions.

@@

#### `gen_scurve()`

To simulate an S-curve structure in a higher-dimensional space, we define the function `gen_scurve(n, p)`, which generates $n$ observations in $p\text{-}D$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:scurve) a"} else { "Figure \\@ref(fig:scurve-proj) a"})`).

The $3\text{-}D$ geometry is constructed by introducing a latent parameter, $\theta \sim U\left(-3\pi/2, 3\pi/2\right)$. This parameter controls the curvature of the manifold. The first three dimensions form the S-curve structure: $X_1 = \sin(\theta)$, $X_2 \sim U(0, 2)$, $X_3 = \text{sign}(\theta)(\cos(\theta) - 1)$. This configuration creates a horizontally curled shape in $(X_1, X_3)$, with additional band thickness in the $X_2$ direction. For $p > 3$, additional noise dimensions are appended introducing structured, wavy noise.

@@

#### `gen_scurvehole()`

To simulate a variant of the S-curve structure with a removed region ("hole"), we define the function `gen_scurvehole(n, p)` (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:scurve) b"} else { "Figure \\@ref(fig:scurve-proj) b"})`). This function produces $n$ observations in $p\text{-}D$ where the first three dimensions describe the S-curve manifold, and remaining dimensions add low-variance Gaussian noise. A subset of observations near a designated anchor point is excluded to introduce a hole in the manifold.

To simulate missing regions on the manifold, a fixed anchor point $(0, 1, 0, \ldots)$ is defined in $p\text{-}D$. All observations within a Euclidean distance of $\sqrt{0.3}$ from the anchor are removed if $\sum_{j=1}^p (X_j - a_j)^2 \leq 0.3$. This exclusion creates a hole in the manifold centered near the middle vertical region of the S-curve.

>>>>

An S-curve structure (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:scurve)"} else { "Figure \\@ref(fig:scurve-proj)"})`) simulates data that lies along a smooth, non-linear manifold. The functions generate both the standard S-curve shape and, similar to the cube generator, an S-curve variant with structured hole that introduce missing or incomplete region. These variations are useful for evaluating how well algorithms capture non-linear geometry and handle incomplete manifolds in high-dimensional data. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:scurve-tb-html)', '\\@ref(tab:scurve-tb-pdf)'))` summarizes these functions.

@@

For the S-curve structure, `gen_scurve(n, p)` (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:scurve) a"} else { "Figure \\@ref(fig:scurve-proj) a"})`), the $3\text{-}D$ geometry is constructed by introducing a latent parameter, $\theta \sim U\left(-3\pi/2, 3\pi/2\right)$. This parameter controls the curvature of the manifold. The first three dimensions form the S-curve structure: $X_1 = \sin(\theta)$, $X_2 \sim U(0, 2)$, $X_3 = \text{sign}(\theta)(\cos(\theta) - 1)$. This configuration creates a horizontally curled shape in $(X_1, X_3)$, with additional band thickness in the $X_2$ direction. For $p > 3$, additional noise dimensions are appended to embed the structure in higher dimensions.

@@

To introduce missing or incomplete regions on the manifold, `gen_scurvehole(n, p)` (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:scurve) b"} else { "Figure \\@ref(fig:scurve-proj) b"})`) removes a subset of points near a designated anchor, creating a hole in the S-curve. This produces a hollowed manifold while preserving the overall curvature and structure of the S-curve.

### 2025-09-19 paper-cardinalR.tex
Updated S-curve section

An S-curve structure (Figure \ref{fig:scurve-proj}) simulates data that lies along a smooth, non-linear manifold, making it a classic example for testing the performance of non-linear dimension reduction methods. This package provides functions to generate standard S-curve shapes as well as variants containing holes to introduce missing or incomplete regions. These variations are useful for evaluating how well algorithms capture non-linear geometry and handle incomplete manifolds in high-dimensional data. Table \ref{tab:scurve-tb-pdf} summarizes these functions.

@@

\paragraph{\texorpdfstring{\texttt{gen\_scurve()}}{gen\_scurve()}}\label{gen_scurve}

To simulate an S-curve structure in a higher-dimensional space, we define the function \texttt{gen\_scurve(n,\ p)}, which generates \(n\) observations in \(p\text{-}D\) (Figure \ref{fig:scurve-proj} a).

The \(3\text{-}D\) geometry is constructed by introducing a latent parameter, \(\theta \sim U\left(-3\pi/2, 3\pi/2\right)\). This parameter controls the curvature of the manifold. The first three dimensions form the S-curve structure: \(X_1 = \sin(\theta)\), \(X_2 \sim U(0, 2)\), \(X_3 = \text{sign}(\theta)(\cos(\theta) - 1)\). This configuration creates a horizontally curled shape in \((X_1, X_3)\), with additional band thickness in the \(X_2\) direction. For \(p > 3\), additional noise dimensions are appended introducing structured, wavy noise.

@@

\paragraph{\texorpdfstring{\texttt{gen\_scurvehole()}}{gen\_scurvehole()}}\label{gen_scurvehole}

To simulate a variant of the S-curve structure with a removed region (``hole''), we define the function \texttt{gen\_scurvehole(n,\ p)} (Figure \ref{fig:scurve-proj} b). This function produces \(n\) observations in \(p\text{-}D\) where the first three dimensions describe the S-curve manifold, and remaining dimensions add low-variance Gaussian noise. A subset of observations near a designated anchor point is excluded to introduce a hole in the manifold.

To simulate missing regions on the manifold, a fixed anchor point \((0, 1, 0, \ldots)\) is defined in \(p\text{-}D\). All observations within a Euclidean distance of \(\sqrt{0.3}\) from the anchor are removed if \(\sum_{j=1}^p (X_j - a_j)^2 \leq 0.3\). This exclusion creates a hole in the manifold centered near the middle vertical region of the S-curve.

>>>>

An S-curve structure (Figure \ref{fig:scurve-proj}) simulates data that lies along a smooth, non-linear manifold. The functions generate both the standard S-curve shape and, similar to the cube generator, an S-curve variant with structured hole that introduce missing or incomplete region. These variations are useful for evaluating how well algorithms capture non-linear geometry and handle incomplete manifolds in high-dimensional data. Table \ref{tab:scurve-tb-pdf} summarizes these functions.

@@

For the S-curve structure, \texttt{gen\_scurve(n,\ p)} (Figure \ref{fig:scurve-proj} a), the \(3\text{-}D\) geometry is constructed by introducing a latent parameter, \(\theta \sim U\left(-3\pi/2, 3\pi/2\right)\). This parameter controls the curvature of the manifold. The first three dimensions form the S-curve structure: \(X_1 = \sin(\theta)\), \(X_2 \sim U(0, 2)\), \(X_3 = \text{sign}(\theta)(\cos(\theta) - 1)\). This configuration creates a horizontally curled shape in \((X_1, X_3)\), with additional band thickness in the \(X_2\) direction. For \(p > 3\), additional noise dimensions are appended to embed the structure in higher dimensions.

@@

To introduce missing or incomplete regions on the manifold, \texttt{gen\_scurvehole(n,\ p)} (Figure \ref{fig:scurve-proj} b) removes a subset of points near a designated anchor, creating a hole in the S-curve. This produces a hollowed manifold while preserving the overall curvature and structure of the S-curve.
